### 2020-07-10 paper.tex
Add some text to main paper

\subsection{Traces}

@@

\begin{algorithm}[h]
\begin{algorithmic}
\Procedure{involutive-mcmc-move}{$p$, $q$, $f$, $x$}
    \State $u \sim q(\cdot; x)$ \Comment{Sample auxiliary variables}
    \State $(x', u') \gets f(x, u)$ \Comment{Apply involution}
    %\State $\alpha \gets
        %\frac{p(x') q(u'; x')}{p(x) q(u; x)} \left| \left[ \frac{\partial h(u, t)}{\partial (u, t)} \right] \right|$
    % TODO or should we show the Radon-Nikodym derivative?
    \State $\alpha \gets
        \displaystyle \frac{p(x') q_{x'}(u')}{p(x) q_{x}(u)} \left( \frac{d (\mu \circ f^{-1})}{d \mu} (x, u)\right)$
    % TODO check it
    \State $r \sim \mathrm{Uniform}(0, 1)$
    \State \algorithmicif \, $r \le \alpha$ \algorithmicthen \, \Return $x'$ \algorithmicelse \, \Return $x$ 
\EndProcedure
\end{algorithmic}
\caption{Involutive MCMC}
\label{alg:involutive-mcmc}
\end{algorithm}

\begin{lemma}\label{lemma:involution-detailed-balance} % TODO cite
For $p$, $q$, and $f$ satisfying the properties above, let $k'$ denote the following transition kernel on $Z$:
\[
k'_z(A) := [f(z) \in A] \alpha(z) + [z \in A](1 - \alpha(z))
\]
The transition kernel $k'$ on $Z$ defined by the involution $f$ satisfies detailed balance with respect to $\pi$:
\[
\int_B k'_z(A) \pi(dz) = \int_A k'_z(B) \pi(dz) \mbox{ for all } A, B \in \Sigma_Z
\]
And in particular it is stationary with respect to $\pi$:
\[
\int_Z k'_z(A) \pi(dz) = \pi(A) \mbox{ for all } A \in \Sigma_Z
\]
\end{lemma}
\begin{proof}
\end{proof}

\begin{theorem}\label{theorem:stationarity}
The transition kernel $k$ on $X$ defined by Algorithm~\ref{alg:involutive-mcmc} for valid $p$, $q$, and $f$:
\[
k_x(A) := \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \;\; \mbox{for all} \;\; A \in \Sigma_X, x \in X
\]
is stationary with respect to $p$.
\end{theorem}
\begin{proof}
By Lemma~\citep{lemma:involution-detailed-balance}, the transition kernel $k'$ is stationary with respect to $\pi$:
\[
%\int_B k'_{z}(B) \pi(z) \mu(dz) = \pi(B) \mbox{ for all } z \in X \times U, B \in \Sigma_Z
\]
It suffices to show that
\begin{align*}
\int_X k_x(A) p(x) \mu_X(dx) &= p(A) \mbox{ for all } A \in \Sigma_X\\
\int_X \left( \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \right) p(x) \mu_X(dx) &= p(A)\\
\int_{X \times U} k'_{x,u}(A \times U) q_x(u) p(x) \mu(dz) &= p(A)
\end{align*}
By Lemma~\citep{lemma:involution-detailed-balance}, setting $B := X \times U$, the left-hand-side is:
\[
\pi(X \times U) = \int_X \int_U p(x) q_x(u) \mu(
\]
\end{proof}
\paragraph{Detailed balance of the involution}
%Let $q_

\paragraph{Stationarity}

\paragraph{Conditional distribution and observations fixed}

@@

\subsection{Distributions on traces}
% TODO rewrite this section..
Define target density $\pi$ (XXX below it is called $p$, fix this) on the joint space $X \times U$, where $X$ are traces of the model and $X$ are traces of the proposal, in terms of the density $p(x)$ and the family of densities $q_x(u)$ as $\pi(x, u) := p(x) q_u(x)$.

% TODO rename the joint target density to \pi below
We want to show stationarity with respect to $\pi$, using stationarity of the involution with respect to the joint density (below).
\begin{equation}
\int_X p(x) \ell_x(A) \mu(dx) = p(A) \mbox{ for all } A \in \Sigma
\end{equation}
where $\ell_x(A)$ is the probability of transitioning into set $A$ from $x$, and is given by:
\begin{equation}
\ell_x(A) := \int_U k_{(x, u)}(A \times U) \mu(du)
\end{equation}
where $k_{(x, u)}$ is the measure on $X \times U$ that is induced by applying the involution $h$ to the state $(x, u)$.
% TODO checkme

\subsection{Derivation of the pushforward Radon-Nikodym derivative for a special case}
We consider the special case in which the involution $f$ can be factored into an involution on a countable set $I$ and a family of bijections on $\mathbb{R}^{n_i}$ for some $n_i$ for each $i \in I$.
Suppose $Z = \{(i, \mathbf{x}) : i \in I, \mathbf{x} \in \mathbb{R}^{n_i}\}$.
Suppose $f_1$ is an involution on $I$ and $n_i = n_{f_1(i)}$ and $f_2$ is a family of continuously differentiable bijections indexed by $i \in I$, such that $f_{2,i} : \mathbb{R}^{n_i} \to \mathbb{R}^{n_i}$.
Also suppose that $f_{2,i} = f_{2,f_1(i)}^{-1}$ for all $i \in I$. That is,
\begin{equation}
f_{2,i}(f_{2,f_1(i)}(\mathbf{x})) = \mathbf{x} \mbox{ for all } \mathbf{x} \in \mathbb{R}^{n_i} \mbox{ and all } i \in I
\end{equation} 
Then, $f : Z \to Z : (i, \mathbf{x}) \mapsto (f_1(i), f_{2,i}(\mathbf{x}))$ is an involution because:
\begin{equation}
f(f(i, \mathbf{x})) = f(f_1(i), f_{2,i}(\mathbf{x})) = (f_1(f_1(i)), f_{2,f_1(i)}(f_{2,i}(\mathbf{x}))) = (i, \mathbf{x})
\end{equation}
Let $\Sigma_n$ and $\mu_n$ denote the Lebesgue $\sigma$-algebra and Lebesgue measure on $\mathbb{R}^n$, respectively.
Let $\Sigma \subset \mathcal{P}(Z)$ be the $\sigma$-algebra of sets of the form $\cup_{i \in I} \{(i, \mathbf{x}) : \mathbf{x} \in A_i\}$ for some $A_i \in \Sigma_{n_i}$ for each $i \in I$.
%Let $\Sigma$ be the $\sigma$-algebra on $Z$ generated by the sets $G := \{ \{(i, \mathbf{x}) : \mathbf{x} \in A\} : i \in I, A \in \Sigma_{n_i} \}$.
%Note that the sets $G$ form a $\pi$-system:
%\begin{align}
%\{(i, \mathbf{x}) : \mathbf{x} \in A\} \cap \{(j, \mathbf{x}) : \mathbf{x} \in B\} = \varnothing = \{(i, \mathbf{x}) : \mathbf{x} \in \varnothing \} \in G \mbox{ for } i \ne j\\
%\{(i, \mathbf{x}) : \mathbf{x} \in A\} \cap \{(i, \mathbf{x}) : \mathbf{x} \in B\} = \{(i, \mathbf{x}) : \mathbf{x} \in A \cap B\} \in G \mbox{ because } A \cap B \in \Sigma_{n_i}
%\end{align}
Let $\mu$ denote the measure on measurable space $(Z, \Sigma)$ given by:%, uniquely defined by its value on the sets $G$:
\begin{equation}
\mu(\cup_{i \in I} \{(i, \mathbf{x}) : \mathbf{x} \in A_i\}) := \sum_{i \in I} \mu_{n_i}(A_i)
\end{equation}
We wish to show that the Radon-Nikodym derivative of the pushforward of $\mu$ by $f$ with respect to $\mu$, evaluated at $(i, \mathbf{x})$, is the absolute value of the Jacobian (determinant) of the function $f_{2,i}$ evaluated at $\mathbf{x}$, which is denoted $(J f_{2,i})(\mathbf{x})$:
\begin{equation}
\frac{d (\mu \circ f^{-1})}{d \mu}(i, \mathbf{x})
= \left| (Jf_{2,i})(\mathbf{x})\right|
\end{equation}
Consider $(\mu \circ f^{-1})(A)$ for $A \in \Sigma$:
\begin{align}
(\mu \circ f^{-1})(\cup_{i \in I} \{ (i, \mathbf{x}) : \mathbf{x} \in A_i \})
&= \mu(f^{-1}(\cup_{i \in I} \{ (i, \mathbf{x}) : \mathbf{x} \in A_i \}))\\
&= \mu(\cup_{i \in I} f^{-1}(\{ (i, \mathbf{x}) : \mathbf{x} \in A_i \}))\\
&= \mu(\cup_{i \in I} \{ (f_1(i), \mathbf{x}) : \mathbf{x} \in f_{2,i}(A_i) \})\\
&= \sum_{i \in I} \mu_{n_i}(f_{2,i}(A_i))
\end{align}
It suffices to show that for all $A \in \Sigma$:
\begin{equation}
\int_A \left| (Jf_{2,i})(\mathbf{x})\right| \mu(dz) = \sum_{i \in I} \mu_{n_i}(f_{2,i}(A_i))
\end{equation}
Expanding the left-hand side:
\begin{align}
\int_A \left| (Jf_{2,i})(\mathbf{x})\right| \mu(dz)
= \sum_{i \in I} \int_{A_i} \left| (Jf_{2,i})(\mathbf{x})\right| \mu_{n_i}(d \mathbf{x}) % TODO how to justify this? it seems obvious
= \sum_{i \in I} \mu_{n_i}(f_{2,i}(A_i)) % Borgachev Theorem 4.7.1 
\end{align}
where the final step uses Borgachev Theorem 3.7.1 with $g := 1$, and $F := f_{2,i}$.

\subsection{Proof of detailed balance for involution}
\citet{tierney1998note} gives a class of MCMC kernels based on involutions that satisfy detailed balance.
Specifically, given an involution $T : Z \to Z$, they denote the pushforward of $\pi$ under $T$ by $\pi'$:
$\pi'(A) := \pi(T^{-1}(A))$.
They let $\nu(A) := \pi(A) + \pi'(A)$.
They let $h(z)$ denote `a density for $\pi$ with respect to $\nu$'.
Then, they set:
\[
\alpha(x \to T(x)) \frac{h(x)}{h(T(x))} = \alpha(T(x) \to x)
\]
Our definition of $\alpha$ satisfies:
\[
\frac{\alpha(z \to T(z))}{\alpha(T(z) \to z)} = \frac{\pi(T(z))}{\pi(z)} \left( \frac{d (\mu \circ T^{-1})}{d \mu} (z) \right)
\]
Therefore, to use Tierney's result, it suffices to show that (when $\pi$ and $\pi'$ are $\sigma$-finite and have densities with respect to the same reference measure $\mu$):
\[
\frac{h(T(z))}{h(z)}= \frac{\pi(T(z))}{\pi(z)} \left( \frac{d (\mu \circ T^{-1})}{d \mu} (z) \right)
\]
(note that $T$ is our $f$).
What is $h(z)$, the density of $\pi$ with respect to $\nu(A) := \pi(A) + \pi'(A)$?
Well, if $\pi$ and $\pi'$ have densities with respect to $\mu$, then so does $\nu$, and it is $\nu(z) = \pi(z) + \pi'(z)$, so the density of $\pi$ with respect to $\nu$ is:
\[
h(z) = \frac{\pi(z)}{\nu(z)} %= \frac{\pi(z)}{\pi(z) + \pi'(z)}
\]
and therefore:
\[
\frac{h(T(z))}{h(z)} = \frac{\pi(T(z))}{\pi(z)} \cdot \frac{\nu(z)}{\nu(T(z))}
\]
It suffices to show that:
\[
\frac{\nu(T(z))}{\nu(z)} = \left( \frac{d (\mu \circ T^{-1})}{d \mu} \right)^{-1}
\]
First, note that $\nu$ and $\nu \circ T^{-1}$ are the same measure:
\begin{align}
\nu(A) &= \pi(A) + \pi'(A) = \pi(A) + \pi(T^{-1}(A))\\
(\nu \circ T^{-1})(A) &= \pi(T^{-1}(A)) + \pi(T^{-1}(T^{-1}(A))) = \pi(T^{-1}(A)) + \pi(A)
\end{align}
Therefore,
\[
\frac{d (\nu \circ T^{-1})}{d \nu}(z) = 1 \mbox{ for all } z
\]
Expanding $d(\nu \circ T^{-1}) / d \nu$ using the chain rule:
\begin{align}
1 &= \frac{d (\nu \circ T^{-1})}{d \nu}(z)\\
&=   \frac{d (\nu \circ T^{-1})}{d (\mu \circ T^{-1})}(z) \cdot
    \frac{d (\mu \circ T^{-1})}{d \mu}(z) \cdot
    \frac{d \mu}{d \nu}(z)\\
&= \nu(T(z))  \frac{d (\mu \circ T^{-1})}{d \mu} \frac{1}{\nu(z)}
\end{align}
where in the second step we used that
\begin{equation}
\frac{d (\nu \circ T^{-1})}{d (\mu \circ T^{-1})}(z) = \frac{d \nu}{d \mu}(T(z)) = \nu(T(z))
\end{equation}
To show this, we use the definition of the Radon-Nikodym derivative $d (\nu \circ T^{-1}) / d (\mu \circ T^{-1})$.
To show that $z \mapsto (d \nu / d \mu)(T(z))$ is this Radon-Nikodym derivative, it suffices to show that for all measurable $A$:
\begin{equation}
\int_A \left( \frac{d \nu}{d \mu} (T(z)) \right) ( \mu \circ T^{-1} )(dz) = (\nu \circ T^{-1})(A) := \nu(T^{-1}(A))
\end{equation}
Applying Theorem 3.6.1 in Bogachev with $Y := A$, $y := z$, $x := z'$, $X := T^{-1}(A)$, $f := T$, and $g(y) := (d\nu / d\mu)(T(y))$:
\begin{align}
\int_A \left( \frac{d \nu}{d \mu} (T(z)) \right) ( \mu \circ T^{-1} )(dz)
&= \int_Y g(y) ( \mu \circ f^{-1} )(dy)\\
&= \int_X g(f(x)) \mu(dx) \;\;\;\; [\mbox{Bogachev Theorem 3.6.1}]\\
&= \int_{T^{-1}(A)} (d\nu / d\mu)(T(T(z'))) \mu(dz')\\
&= \int_{T^{-1}(A)} (d\nu / d\mu)(z') \mu(dz')\\
&= \int_{T^{-1}(A)} \nu(dz')
= \nu(T^{-1}(A))
\end{align}
% NOTE Bogachev's theorem uses domains of integration which are the whole
% sample spaces for the two sigma algebras.  I think that the theorem should
% also apply to any measurable sets A and T(A) because one can construct a
% sigma algebra and measure that are restricted to to any nonempty measurable
% set.

\subsection{Proof of stationarity for involution}
Stationarity follows from detailed balance by setting $B := Z$ in the detailed balance equation, and using $k'_z(Z) = \alpha(z) + (1 - \alpha(z)) =1$, so the right-hand side simplifies to $\int_A \pi(dz) = \pi(A)$.

\subsection{Proof of stationarity for end-to-end kernel}
We are given that the involution is stationary with respect to (the measure induced by) $\pi(x, u) := p(x) q_x(u)$:
\[
\int_Z k'_z(A) \pi(z) \mu(dz) = \pi(A) \mbox{ for all } A \in \Sigma_Z
\]
It suffices to show that the end-to-end kernel $k$ is stationary with respect to (the measure induced by) $p(x)$:
\[
\int_X k_x(A) p(x) \mu_X(dx) = p(A) \mbox{ for all } A \in \Sigma_X
\]
The kernel $k$ on $X$ defined by Algorithm~\ref{alg:involutive-mcmc} for valid $p$, $q$, and $f$:
\[
k_x(A) := \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \;\; \mbox{for all} \;\; A \in \Sigma_X, x \in X
\]
is stationary with respect to $p$.
\begin{align*}
    \int_X k_x(A) p(x) \mu_X(dx) &= p(A) \mbox{ for all } A \in \Sigma_X\\
    \int_X \left( \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \right) p(x) \mu_X(dx) &= p(A)\\
    \int_{X \times U} k'_{x,u}(A \times U) q_x(u) p(x) \mu(dz) &= p(A)\\
    \int_{X \times U} k'_{x,u}(A \times U) \pi(z) \mu(dz) &= p(A)\\
    \pi(A \times U) &= p(A)\\
    \int_{A \times U} q_x(u) p(x) \mu(dz) &= p(A)\\
    \int_A \left( \int_U q_x(u) \mu_U(du) \right) p(x) \mu_X(dx) &= p(A)\\
    \int_A p(x) \mu_X(dx) &= p(A)
\end{align*}
% TODO characterize the various assumptions that allow us to e.g. rearrange the order of integration, etc.

>>>>

%\subsection{Traces}

@@

\begin{algorithm}[h]
\begin{algorithmic}
\Procedure{involutive-mcmc-move}{$p$, $q$, $f$, $x$}
    \State $u \sim q(\cdot; x)$ \Comment{Sample auxiliary variables}
    \State $(x', u') \gets f(x, u)$ \Comment{Apply involution}
    %\State $\alpha \gets
        %\frac{p(x') q(u'; x')}{p(x) q(u; x)} \left| \left[ \frac{\partial h(u, t)}{\partial (u, t)} \right] \right|$
    % TODO or should we show the Radon-Nikodym derivative?
    \State $\alpha \gets
        \displaystyle \frac{p(x') q_{x'}(u')}{p(x) q_{x}(u)} \left( \frac{d (\mu \circ f^{-1})}{d \mu} (x, u)\right)$
    % TODO check it
    \State $r \sim \mathrm{Uniform}(0, 1)$
    \State \algorithmicif \, $r \le \alpha$ \algorithmicthen \, \Return $x'$ \algorithmicelse \, \Return $x$ 
\EndProcedure
\end{algorithmic}
\caption{Involutive MCMC}
\label{alg:involutive-mcmc}
\end{algorithm}
See Section~\ref{sec:radon-nikodym-special-case} for a derivation of the Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d \mu$ for a special case, when the involution $f$ can be factored into (i) an involution on a countable partition of $Z$ and (ii) a family of continuously differentiable bijection on Euclidean spaces.

\begin{theorem}[Involutive MCMC is stationary]
For $p$, $q$, and $f$ satisfying the requirements above, Algorithm~\ref{alg:involutive-mcmc} defines a probability kernel on $X$ that is stationary with respect to the distribution $p$.
\end{theorem}
\begin{proof}
The proof is presented in stages in the appendix (see Section~\ref{sec:involution-detailed-balance}, Section~\ref{sec:involution-is-stationary}, and Section~\ref{sec:involutive-mcmc-is-stationary}).
\end{proof}

%\begin{lemma}\label{lemma:involution-detailed-balance} % TODO cite
%For $p$, $q$, and $f$ satisfying the properties above, let $k'$ denote the following transition kernel on $Z$:
%\[
%k'_z(A) := [f(z) \in A] \alpha(z) + [z \in A](1 - \alpha(z))
%\]
%The transition kernel $k'$ on $Z$ defined by the involution $f$ satisfies detailed balance with respect to $\pi$:
%\[
%\int_B k'_z(A) \pi(dz) = \int_A k'_z(B) \pi(dz) \mbox{ for all } A, B \in \Sigma_Z
%\]
%And in particular it is stationary with respect to $\pi$:
%\[
%\int_Z k'_z(A) \pi(dz) = \pi(A) \mbox{ for all } A \in \Sigma_Z
%\]
%\end{lemma}
%\begin{proof}
%\end{proof}
%
%\begin{theorem}\label{theorem:stationarity}
%The transition kernel $k$ on $X$ defined by Algorithm~\ref{alg:involutive-mcmc} for valid $p$, $q$, and $f$:
%\[
%k_x(A) := \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \;\; \mbox{for all} \;\; A \in \Sigma_X, x \in X
%\]
%is stationary with respect to $p$.
%\end{theorem}
%\begin{proof}
%By Lemma~\citep{lemma:involution-detailed-balance}, the transition kernel $k'$ is stationary with respect to $\pi$:
%\[
%%\int_B k'_{z}(B) \pi(z) \mu(dz) = \pi(B) \mbox{ for all } z \in X \times U, B \in \Sigma_Z
%\]
%It suffices to show that
%\begin{align*}
%\int_X k_x(A) p(x) \mu_X(dx) &= p(A) \mbox{ for all } A \in \Sigma_X\\
%\int_X \left( \int_U k'_{x,u}(A \times U) q_x(u) \mu_U(du) \right) p(x) \mu_X(dx) &= p(A)\\
%\int_{X \times U} k'_{x,u}(A \times U) q_x(u) p(x) \mu(dz) &= p(A)
%\end{align*}
%By Lemma~\citep{lemma:involution-detailed-balance}, setting $B := X \times U$, the left-hand-side is:
%\[
%\pi(X \times U) = \int_X \int_U p(x) q_x(u) \mu(
%\]
%\end{proof}
%\paragraph{Detailed balance of the involution}
%%Let $q_
%
%\paragraph{Stationarity}
%
%\paragraph{Conditional distribution and observations fixed}

@@

%\subsection{Distributions on traces}
%% TODO rewrite this section..
%Define target density $\pi$ (XXX below it is called $p$, fix this) on the joint space $X \times U$, where $X$ are traces of the model and $X$ are traces of the proposal, in terms of the density $p(x)$ and the family of densities $q_x(u)$ as $\pi(x, u) := p(x) q_u(x)$.
%
%% TODO rename the joint target density to \pi below
%We want to show stationarity with respect to $\pi$, using stationarity of the involution with respect to the joint density (below).
%\begin{equation}
%\int_X p(x) \ell_x(A) \mu(dx) = p(A) \mbox{ for all } A \in \Sigma
%\end{equation}
%where $\ell_x(A)$ is the probability of transitioning into set $A$ from $x$, and is given by:
%\begin{equation}
%\ell_x(A) := \int_U k_{(x, u)}(A \times U) \mu(du)
%\end{equation}
%where $k_{(x, u)}$ is the measure on $X \times U$ that is induced by applying the involution $h$ to the state $(x, u)$.
% TODO checkme

\subsection{Derivation of the pushforward Radon-Nikodym derivative for a special case} \label{sec:radon-nikodym-special-case}
Implementing Algorithm~\ref{alg:involutive-mcmc} requires computing the Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d \mu$.
This section derives that function for the special case in which the involution $f$ can be factored into an involution on a countable set $I$ and a family of bijections on $\mathbb{R}^{n_i}$ for some $n_i$ for each $i \in I$.
Suppose $Z = \{(i, \mathbf{x}) : i \in I, \mathbf{x} \in \mathbb{R}^{n_i}\}$.
Suppose $f_1$ is an involution on $I$ and $n_i = n_{f_1(i)}$ and $f_2$ is a family of continuously differentiable bijections indexed by $i \in I$, such that $f_{2,i} : \mathbb{R}^{n_i} \to \mathbb{R}^{n_i}$.
Also suppose that $f_{2,i} = f_{2,f_1(i)}^{-1}$ for all $i \in I$. That is,
\begin{equation}
f_{2,i}(f_{2,f_1(i)}(\mathbf{x})) = \mathbf{x} \mbox{ for all } \mathbf{x} \in \mathbb{R}^{n_i} \mbox{ and all } i \in I
\end{equation} 
Then, $f : Z \to Z : (i, \mathbf{x}) \mapsto (f_1(i), f_{2,i}(\mathbf{x}))$ is an involution because:
\begin{equation}
f(f(i, \mathbf{x})) = f(f_1(i), f_{2,i}(\mathbf{x})) = (f_1(f_1(i)), f_{2,f_1(i)}(f_{2,i}(\mathbf{x}))) = (i, \mathbf{x})
\end{equation}
Let $\Sigma_n$ and $\mu_n$ denote the Lebesgue $\sigma$-algebra and Lebesgue measure on $\mathbb{R}^n$, respectively.
Let $\Sigma \subset \mathcal{P}(Z)$ be the $\sigma$-algebra of sets of the form $\cup_{i \in I} \{(i, \mathbf{x}) : \mathbf{x} \in A_i\}$ for some $A_i \in \Sigma_{n_i}$ for each $i \in I$.
%Let $\Sigma$ be the $\sigma$-algebra on $Z$ generated by the sets $G := \{ \{(i, \mathbf{x}) : \mathbf{x} \in A\} : i \in I, A \in \Sigma_{n_i} \}$.
%Note that the sets $G$ form a $\pi$-system:
%\begin{align}
%\{(i, \mathbf{x}) : \mathbf{x} \in A\} \cap \{(j, \mathbf{x}) : \mathbf{x} \in B\} = \varnothing = \{(i, \mathbf{x}) : \mathbf{x} \in \varnothing \} \in G \mbox{ for } i \ne j\\
%\{(i, \mathbf{x}) : \mathbf{x} \in A\} \cap \{(i, \mathbf{x}) : \mathbf{x} \in B\} = \{(i, \mathbf{x}) : \mathbf{x} \in A \cap B\} \in G \mbox{ because } A \cap B \in \Sigma_{n_i}
%\end{align}
Let $\mu$ denote the measure on measurable space $(Z, \Sigma)$ given by:%, uniquely defined by its value on the sets $G$:
\begin{equation}
\mu(\cup_{i \in I} \{(i, \mathbf{x}) : \mathbf{x} \in A_i\}) := \sum_{i \in I} \mu_{n_i}(A_i)
\end{equation}
We wish to show that the Radon-Nikodym derivative of the pushforward of $\mu$ by $f$ with respect to $\mu$, evaluated at $(i, \mathbf{x})$, is the absolute value of the Jacobian (determinant) of the function $f_{2,i}$ evaluated at $\mathbf{x}$, which is denoted $(J f_{2,i})(\mathbf{x})$:
\begin{equation}
\frac{d (\mu \circ f^{-1})}{d \mu}(i, \mathbf{x})
= \left| (Jf_{2,i})(\mathbf{x})\right|
\end{equation}
Consider $(\mu \circ f^{-1})(A)$ for $A \in \Sigma$:
\begin{align}
(\mu \circ f^{-1})(\cup_{i \in I} \{ (i, \mathbf{x}) : \mathbf{x} \in A_i \})
&= \mu(f^{-1}(\cup_{i \in I} \{ (i, \mathbf{x}) : \mathbf{x} \in A_i \}))\\
&= \mu(\cup_{i \in I} f^{-1}(\{ (i, \mathbf{x}) : \mathbf{x} \in A_i \}))\\
&= \mu(\cup_{i \in I} \{ (f_1(i), \mathbf{x}) : \mathbf{x} \in f_{2,i}(A_i) \})\\
&= \sum_{i \in I} \mu_{n_i}(f_{2,i}(A_i))
\end{align}
It suffices to show that for all $A \in \Sigma$:
\begin{equation}
\int_A \left| (Jf_{2,i})(\mathbf{x})\right| \mu(dz) = \sum_{i \in I} \mu_{n_i}(f_{2,i}(A_i))
\end{equation}
Expanding the left-hand side:
\begin{align}
\int_A \left| (Jf_{2,i})(\mathbf{x})\right| \mu(dz)
= \sum_{i \in I} \int_{A_i} \left| (Jf_{2,i})(\mathbf{x})\right| \mu_{n_i}(d \mathbf{x}) % TODO how to justify this? it seems obvious
= \sum_{i \in I} \mu_{n_i}(f_{2,i}(A_i)) % Borgachev Theorem 4.7.1 
\end{align}
where the final step uses Borgachev Theorem 3.7.1 with $g := 1$, and $F := f_{2,i}$.

\subsection{Proof of detailed balance for involution} \label{sec:involution-detailed-balance}
\citet{tierney1998note} gives a class of MCMC kernels based on involutions that satisfy detailed balance.
We reproduce the result in our notation:
\begin{lemma}[Detailed balance for involution move~\citep{tierney1998note}]
Let $(Z, \Sigma, \pi)$ denote a measure space.
Suppose $f$ is a one-to-one function from $Z$ onto $Z$ such that $f^{-1} = f$.
Consider the probability kernel $k$ defined by $k_z(A) := [f(z) \in A] \alpha(z, f(z)) + [z \in A] (1 - \alpha(z, f(z)))$ (where $\alpha(z, f(z))$ gives the probability of accepting a proposed transition from $z$ to $f(z)$).
Let $\nu(dz) := \pi(dz) + (\pi \circ f^{-1})(dz)$.
Let $h(z)$ be a density for $\pi$ with respect to $\nu$.
Let $A := \{z \in Z : h(z) > 0 \mbox{ and } h(f(z)) > 0\}$.
$k$ satisfies detailed balance with respect to $\pi$ if and only if:
\begin{enumerate}
\item $\alpha(z, f(z)) = 0$ for $\pi$-almost all $z \not \in A$
\item $\alpha(z, f(z)) \frac{h(z)}{h(f(z))} = \alpha(f(z), z)$
\end{enumerate}
\end{lemma}

Now we apply Lemma~\ref{lemma:tierney-involution} to our setting where $\pi$ is $\sigma$-finite, there exists a $\sigma$-finite reference measure $\mu$ for measurable space $(Z, \Sigma)$ such that $\pi$ is absolutely continuous with respect to $\mu$, and where the pushforward of $\mu$ by $f$, denoted $\mu \circ f^{-1}$, is absolutely continuous with respect to $\mu$.
(One step in the proof---Equation~\ref{eq:mu-is-absolutely-continuous-with-respect-to-nu}---will also require that $\mu$ is also absolutely continuous with respect to $\pi$, which is possible if we restrict $Z$ to the support of $\pi$, $\{z : (d \pi / d \mu)(z) > 0\}$, which must be in the $\sigma$-algebra $\Sigma$).

In our setting, $\alpha$ is defined as:
\begin{equation}
\alpha(z, f(z)) := \mbox{min}\left\{1, \frac{\frac{d \pi}{d \mu}(f(z))}{\frac{d \pi}{d \mu}(z)} \cdot \frac{d (\mu \circ f^{-1})}{d \mu}(z)\right\}
\end{equation}
This definition of $\alpha$ satisfies:
\begin{equation}
\alpha(z, f(z)) \frac{\frac{d \pi}{d \mu}(z)}{\frac{d \pi}{d \mu}(f(z))} \cdot \left( \frac{d (\mu \circ f^{-1})}{d \mu} (z)\right)^{-1}= \alpha(f(z), z)
%\frac{\alpha(z, f(z))}{\alpha(f(z), z)} = \frac{\pi(T(z))}{\pi(z)} \left( \frac{d (\mu \circ T^{-1})}{d \mu} (z) \right)
\end{equation}
Therefore, to apply Lemma~\ref{lemma:tierney-involution}, it suffices to show $\pi$ has density with respect to $\nu$ (denoted $h(z)$) such that:
\begin{equation}
\frac{h(z)}{h(f(z))} = \frac{\frac{d \pi}{d \mu}(z)}{\frac{d \pi}{d \mu}(f(z))} \left( \frac{d (\mu \circ f^{-1})}{d \mu} (z)\right)^{-1}
\end{equation}
(TODO: be careful with the zero cases and support.)\\ % $ TODO
Since $\pi$ and $\pi \circ f^{-1}$ are both absolutely continuous with respect to $\mu$, $\nu$ is also absolutely continuous with respect to $\mu$, and has density:
\begin{equation}
\frac{d \nu}{d \mu}(z) = \frac{d \pi}{d \mu}(z) + \frac{d (\pi \circ f^{-1})}{d \mu}(z) 
\end{equation}
$\pi$ is absolutely continuous with respect to $\nu$, and therefore:
\begin{equation}
\frac{d \pi}{d \mu}(z) = \frac{d \pi}{d \nu}(z) \cdot \frac{d \nu}{d \mu}(z)
\end{equation}
Therefore,
\begin{equation}
h(z)
:= \frac{d \pi}{d \nu}(z)
= \frac{\frac{d \pi}{d \mu}(z)}{\frac{d \nu}{d \mu}(z)}
\;\; \mbox{ for } z \mbox{ s.t. } (d \nu / d \mu)(z) > 0
\end{equation}
Therefore:%(TODO: be careful with zeros)
\begin{equation}
\frac{h(z)}{h(f(z))} = \frac{\frac{d \pi}{d \mu}(z)}{\frac{d \pi}{d \mu}(f(z))} \cdot \frac{\frac{d \nu}{d \mu}(f(z))}{\frac{d \nu}{d \mu}(z)}
\end{equation}
It suffices to show that:
\[
\frac{\frac{d \nu}{d \mu}(f(z))}{\frac{d \nu}{d \mu}(z)} = \left(\frac{d (\mu \circ f^{-1})}{d \mu}(z)\right)^{-1}
\]

First, we prove a Lemma:
\begin{lemma} \label{lemma:reference-measure-pushforward}
If $(Z, \Sigma)$ is a measurable space and $f : Z \to Z$ is a measurable function that is an involution, $\nu$ and $\mu$ are $\sigma$-finite measures such that $\nu$ is absolutely continuous with respect to $\mu$, and such that the pushforward measures $\nu \circ f^{-1}$ and $\mu \circ f^{-1}$ are both $\sigma$-finite, then
$\nu \circ f^{-1}$ is absolutely continuous with respect to $\mu \circ f^{-1}$ and
\begin{equation}
\frac{d (\nu \circ f^{-1})}{d (\mu \circ f^{-1})}(z) = \frac{d \nu}{d \mu}(f(z))
\end{equation}
\end{lemma}
\begin{proof}
First, $\nu \circ f^{-1}$ is absolutely continuous with respect to $\mu \circ f^{-1}$ because
$(\mu \circ f^{-1})(A) = 0$ implies $\mu(f^{-1}(A)) = 0$ implies $\nu(f^{-1}(A)) = 0$ implies $(\nu \circ f^{-1})(A) = 0$.
To show that $z \mapsto (d \nu / d \mu)(f(z))$ is the Radon-Nikodym derivative $d (\nu \circ f^{-1}) / d (\mu \circ f^{-1})$, it suffices to show that for all $A \in \Sigma$:
\begin{equation}
\int_A \left( \frac{d \nu}{d \mu} (f(z)) \right) ( \mu \circ f^{-1} )(dz) = (\nu \circ f^{-1})(A) := \nu(f^{-1}(A))
\end{equation}
Applying Theorem 3.6.1 in Bogachev with $Y := A$, $y := z$, $x := z'$, $X := f^{-1}(A)$, and $g(y) := (d\nu / d\mu)(f(y))$:
\begin{align}
\int_A \left( \frac{d \nu}{d \mu} (f(z)) \right) ( \mu \circ f^{-1} )(dz)
&= \int_Y g(y) ( \mu \circ f^{-1} )(dy)\\
&= \int_X g(f(x)) \mu(dx) \;\;\;\; [\mbox{Bogachev Theorem 3.6.1}]\\
&= \int_{f^{-1}(A)} (d\nu / d\mu)(f(f(z'))) \mu(dz')\\
&= \int_{f^{-1}(A)} (d\nu / d\mu)(z') \mu(dz')\\
&= \int_{f^{-1}(A)} \nu(dz')
= \nu(f^{-1}(A))
\end{align}
% NOTE Bogachev's theorem uses domains of integration which are the whole
% sample spaces for the two sigma algebras.  I think that the theorem should
% also apply to any measurable sets A and T(A) because one can construct a
% sigma algebra and measure that are restricted to to any nonempty measurable
% set.

\end{proof}
